---
title: "Compositional Data"
output: pdf_document
date: '2022-06-15'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Statistical analysis of metagenomics data

Link: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6459172/pdf/gi-2019-17-1-e6.pdf

* Compositional data is of the form $x = (x_1, \ldots, x_k)$ such that $x_i > 0$ and $\sum_i x_i = T$
* The total $T$ is defined not to be of importance
* Equivalence to some member of the unit simplex $\mathcal{S} = \{x = (x_1, \ldots, x_k); \quad x_i > 0; \quad \sum_i x_i = 1\}$
* Desirata for CoDA
  * Permutation invariance: changing the order of the parts doesn't matter
  * Scale invariance: any member of the equivalence class should have the same results
  * Sub-compositional coherence: results for a subset should be coherent with results for the whole composition
* Log-ratio $\log(x_i / x_j)$
* Log-contrast $\sum_i a_i \log(x_i); \quad \sum_i a_i = 0$
* Additive log-transformation $\text{alr}(x) = (\log(x_i / x_k))_{i < k}$
* Centered log-transformation $\text{clr}(x) = (\log(x_i / g(x)))_i$ where $g(x) = (\prod_i x_i)^{1 / k}$ is the geometric mean
* Isometric log-transformation (not described)
* Microbiome abundance table is a matrix $X$ with $n$ rows (samples) and $k$ columns (taxa) where each entry $x_ij$ provides the number of reads corresponding to taxon $j$ in sample $i$
* Other elements which could be available include the sample data, the taxonomy table, and the phylogenetic tree
* Alpha diversity: within sample diversity
* Richness $R_\text{obs}$ gives the number of species

## Visualizing the Multinomial in the Simplex

Link: http://www.statsathome.com/2017/09/14/visualizing-the-multinomial-in-the-simplex/

* The multinomial is a discretisation of the simplex
* The true value is $(.37, .37, .3)$ but there is no way to estimate that when the sample size is $n = 10$

```{r}
library(tidyverse)
library(driver)
library(combinat)
library(ggtern)

plot_ternary_multinomial_density <- function(n, prob){
  sx <- t(xsimplex(p = 3, n = n))
  dx <- apply(sx, 1, function(x) dmultinom(x, size = n, prob = prob))
  sx <- cbind(as.data.frame(miniclo(sx)), dx)
  colnames(sx) <- c("v1", "v2", "v3", "d")
  
  s <- ggtern(sx, aes(x = v1, y = v2, z = v3)) +
    geom_point(alpha = 0.4, aes(size = d/max(d))) +
    theme_bw() +
    guides(size = FALSE) +
    scale_size(range = c(.1, 2))
  s
}
```

```{r}
prob <- miniclo(c(.37, .37, .3))
d <- setNames(as.data.frame(prob), c("v1", "v2", "v3"))
plot_ternary_multinomial_density(10, prob) +
  geom_point(data = d, color = "red", size = 2, fill = "red")
```
* As $n$ gets larger it's easier to estimate $p$

```{r}
d <- setNames(as.data.frame(prob), c("v1", "v2", "v3"))
plot_ternary_multinomial_density(100, prob) +
  geom_point(data = d, color = "red", size = 2, fill = "red")
```
* As $p$ gets smaller, it's harder to estimate

```{r}
prob <- miniclo(c(.9, .091, .005))
p1 <- plot_ternary_multinomial_density(100, prob) +
  geom_point(data=setNames(as.data.frame(prob), c("v1", "v2", "v3")), color = "red", size = 2, fill="red")

p2 <- p1 + theme_zoom_L(0.2)

grid.arrange(p1, p2, ncol=2)
```
