---
title: "Simulating metagenomic time series data"
author: "Adam Howes"
bibliography: citations.bib
output:
  html_document:
    toc: yes
    df_print: paged
    code_folding: show
abstract: |
  In this document we aim to 1) build a modular understanding of the data generating process for metagenomic data at varying levels of complexity 2) simulate metagenomic time series data for a baseline (no threat) regime and an exponential (threat) regime
---

```{r echo = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  out.width = "95%",
  fig.align = 'center'
)

cbpalette <- c("#56B4E9", "#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

library(tidyverse)
```

## Introduction via simple version

```{r}
n_organism <- 10
baseline_indices <- 2:n_organism
exponential_indices <- 1
n_bp <- 100
k <- 40
n_kmer <- n_bp - k
time_window <- 14
baseline <- 100
exp_initial <- 1
r <- 1
read_depth <- 10^4
```

Suppose that there are `r n_organism` organisms in total in the sample, and nothing else.
Each organism has a genome that is exactly `r n_bp` base-pairs long.
Each genome is broken into $k$-mers of length `r k`.
There are `r n_kmer` $k$-mers from each organism, obtained by subtracting the $k$-mer length from the genome size.
Assume that each $k$-mer is unique, and there are no sequencing errors.
We collect data over a period of `r time_window` days $t = 0, 1, \dots$ at a single environmental monitoring site.

### Organism-level baseline and exponential regimes

We consider two regimes 1) baseline and 2) exponential growth.
In the baseline regime, a species has a concentration $c^{(b)}$ in water of `r baseline` copy per $\mu$L^[The units don't matter much.], independent of the day.
$$
c^{(b)}_t = c^{(b)}
$$
In the exponential growth regime, a species starts at a concentration $c^{(e)}_0$, and over the `r time_window` day window its concentration increases exponentially according to
$$
c^{(e)}_t = c^{(e)}_0 \exp(rt)
$$
where $r$ is the growth rate which we take to be `r r`, and $c^{(e)}_0$ is the initial exponential concentration which we take to be `r exp_initial` -- lower than the baseline concentration as we assume the exponentially increasing organism to be novel.

```{r}
#' The concentration of the baseline organisms over the time window
conc_baseline <- rep(baseline, time_window)
conc_baseline

#' The concentration of the exponentially increasing organism over the time window
conc_exponential <- exp_initial * exp(r * 0:13)
conc_exponential
```

Suppose that organisms `r baseline_indices` follow the baseline behaviour and organism `r exponential_indices` follows the exponential behaviour.
We will represent the true concentrations of each organism with a matrix `C` as follows:

```{r}
C <- matrix(
  data = NA,
  nrow = time_window,
  ncol = n_organism
)

C[, exponential_indices] <- conc_exponential
C[, baseline_indices] <- conc_baseline
C
```

### Sampling $k$-mers

We assume that the true concentration of each $k$-mer is that of the corresponding organism.
This may be represented by copying each column of the matrix `C` `r n_bp` times.

```{r}
K <- matrix(rep(as.numeric(t(C)), each = n_kmer), nrow = time_window, byrow = TRUE)
```

The matrix `K` now has `r nrow(K)` rows, one for each day, and `r ncol(K)` columns, one for each $k$-mer (of which there are `r n_kmer` times `r n_organism`).
We can calculate proportions, where the total number of $k$-mers is given by the row sums:

```{r}
K_norm <- t(apply(K, 1, function(x) x / sum(x), simplify = TRUE))
# useful::topleft(K_norm)
# useful::bottomleft(K_norm)
```

One way to represent the sequencing process is as a sample from the collection of $k$-mers.
For example, we could consider a multinomial sample with probabilities of sampling each $k$-mer given by `K_norm` and sample size given by the read depth `r read_depth`.

To demonstrate this, suppose we simulate the sequencing process on day 1.
The proportions of each $k$-mer are given by `K_norm[1, ]`, and we may sample using `rmultinom`.
A histogram of the $k$-mer counts at day 1 under each regime, showing that the exponential regime is initialised at low count numbers, is given by:

```{r}
sample_one <- rmultinom(1, read_depth, K_norm[1, ])

get_regime <- function(organism_index) {
  if(organism_index %in% baseline_indices) return("baseline")
  if(organism_index %in% exponential_indices) return("exponential")
  else return(NA)
}

#' Testing that this function works as intended
purrr::map_chr(1:11, get_regime)

data.frame(count = sample_one) %>%
  mutate(
    organism = rep(1:n_organism, each = n_kmer),
    regime = str_to_title(purrr::map_chr(organism, get_regime)),
  ) %>%
  ggplot(aes(x = count, group = regime, fill = regime)) +
    geom_histogram(alpha = 0.8) +
    labs(x = "k-mer count", y = "Occurances", fill = "Regime", title = "k-mer counts at day 1") +
    scale_fill_manual(values = cbpalette) +
    theme_minimal()
```

Now, we will take multinomial samples from all of the days with a call to `apply`:

```{r}
sample <- apply(K_norm, 1, function(row) rmultinom(1, read_depth, row))

colnames(sample) <- paste0("day", 1:ncol(sample))
rownames(sample) <- paste0(1:nrow(sample))

sample_df <- sample %>%
  as.data.frame() %>%
  tibble::rownames_to_column("id") %>%
  pivot_longer(
    cols = starts_with("day"),
    names_to = "day",
    values_to = "count",
    names_prefix = "day"
  ) %>%
  mutate(
    id = as.numeric(id),
    day = as.numeric(day),
    kmer = rep(rep(1:n_kmer, each = time_window), times = n_organism),
    organism = rep(1:n_organism, each = n_kmer * time_window)
  )
```

The data frame `sample_df` filtered to the first $k$-mer from the first organism is given by:

`r filter(sample_df, kmer == 1, organism == 1)`

Let's plot the data from organism `r exponential_indices` which we have set to be exponentially increasing:

```{r}
sample_summary <- sample_df %>%
  filter(organism %in% exponential_indices) %>%
  group_by(day) %>%
  summarise(
    count_upper = quantile(count, 0.95),
    count_median = median(count),
    count_lower = quantile(count, 0.05)
  )

ggplot(sample_summary, aes(x = day, ymin = count_lower, y = count_median, ymax = count_upper)) +
    geom_ribbon(alpha = 0.1) +
    geom_point() +
    geom_point(data = filter(sample_df, organism == 1), aes(x = day, y = count, col = kmer),
               alpha = 0.1, inherit.aes = FALSE) + 
    theme_minimal() +
    labs(x = "Day", y = "Number of reads in sample", col = "k-mer")
```

```{r}
pickout_day <- 10
```

With these settings, on day `r pickout_day` the median count of organism 1 is `r filter(sample_summary, day == pickout_day)$count_median`.
At this stage, the exponential growth curve has started to level off because $k$-mers from organism 1 already represent `r 100 * filter(sample_summary, day == pickout_day)$count_median * n_kmer / read_depth`\% of the total reads.
This is unrealistic as it would be unlikely, or for the novel pathogens we are considering essentially impossible, for any one organism to saturate the space.
<!-- * we expect the exponentially increasing viruses we are catching to have a very small initial relative abundance; so you could start this one with a much lower concentration, and then it won’t saturate over the simulation period -->

Now, we will start to make this simulation more realistic.

## Adding noise to the true abundances

In reality, the organism abundances will vary, neither being fixed to a constant or increasing deterministically.
We will now suppose that the abundances vary stochastically as follows.

### Baseline

For the baseline concentration, we will use a lognormal geometric random walk such that
$$
c^{(b)}_t = c^{(b)}_{t - 1} \times \exp \left( \epsilon^{(b)}_t \right)
$$
where $\exp(\epsilon^{(b)}_t) \sim \mathcal{N}(0, \sigma_b^2)$.
Another way to calculate the baseline concentration at time $t$, $c^{(b)}_t$, is as
$$
c^{(b)}_t = c^{(b)}_{0} \times \exp \left( \sum_{s = 1}^t \epsilon^{(b)}_s \right)
$$
Note that the expected value of the cumulative sum of innovations is
$$
\mathbb{E} \left[ \sum_{s = 1}^t \epsilon^{(b)}_s \right] = \sum_{s = 1}^t \mathbb{E} \left[ \epsilon^{(b)}_s \right] = 0,
$$
and the variance is
$$
\mathbb{Var} \left[ \sum_{s = 1}^t \epsilon^{(b)}_s \right] = \sum_{s = 1}^t \mathbb{Var} \left[ \epsilon^{(b)}_s \right] = t\sigma_b^2.
$$

We can use the formula for the moment generating function of a Gaussian random variable $X \sim \mathcal{N}(\mu, \sigma^2)$
$$
m_X(u) = \mathbb{E} \left[ e^{uX} \right] = \exp(\mu u + \sigma^2u^2 / 2)
$$
to know that the expected concentration at time $T$, say, under this model is
$$
\mathbb{E} \left[ c^{(b)}_t \right] = c^{(b)}_{0} \exp(\sigma_b^2/2) > c^{(b)}_{0}.
$$

```{r}
n_baseline_paths <- 8
```

To show how this behaviour looks, let's simulate `r n_baseline_paths` baseline paths.

```{r}
baseline_df <- data.frame(
  day = rep(1:time_window, times = n_baseline_paths), 
  epsilon = rnorm(time_window * n_baseline_paths),
  path_number = as.factor(rep(1:n_baseline_paths, each = time_window))
  ) %>%
  group_by(path_number) %>%
  arrange(path_number) %>%
  mutate(
    cumsum_epsilon = cumsum(epsilon),
    conc = baseline * exp(cumsum_epsilon)
  ) %>%
  ungroup()

baseline_df %>%
  pivot_longer(
    cols = c("epsilon", "cumsum_epsilon", "conc"),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    variable = recode_factor(variable, 
      "epislon" = "epsilon", 
      "cumsum_epsilon" = "cumsum(epsilon)",
      "conc" = "Concentration")
  ) %>%
  ggplot(aes(x = day, y = value, group = path_number, col = path_number)) +
    geom_line() +
    facet_wrap(~variable, ncol = 1, scales = "free") +
    scale_colour_manual(values = cbpalette) +
    theme_minimal() +
    labs(x = "Day", y = "", col = "Path number", title = "Baseline behaviour")
```

The key takeaway for me is the even though the noise is IID and Gaussian, when you take the cumulative sum and exponentiate it can lead to large deviations in concentration.
For example, the maximum concentration value observed was `r max(baseline_df$conc)` -- which is `r max(baseline_df$conc) / baseline` greater than the baseline concentration.

### Exponential

For the exponential regime, we also suppose a geometric lognormal random walk
$$
c^{(e)}_t = c^{(e)}_{t - 1} \times \exp \left( \epsilon^{(e)}_t \right)
$$
where $\exp(\epsilon^{(e)}_t) \sim \mathcal{N}(r, \sigma_e^2)$ and the growth rate $r > 0$.
The expected concentration is $c^{(b)}_{0} \times \exp(rt)$.
Let's simulate some paths from this distribution as before.

```{r}
exponential_df <- data.frame(
  day = rep(1:time_window, times = n_baseline_paths), 
  epsilon = rnorm(time_window * n_baseline_paths, mean = r),
  path_number = as.factor(rep(1:n_baseline_paths, each = time_window))
  ) %>%
  group_by(path_number) %>%
  arrange(path_number) %>%
  mutate(
    cumsum_epsilon = cumsum(epsilon),
    conc = baseline * exp(cumsum_epsilon)
  ) %>%
  ungroup()

exponential_df %>%
  pivot_longer(
    cols = c("epsilon", "cumsum_epsilon", "conc"),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    variable = recode_factor(variable, 
      "epislon" = "epsilon", 
      "cumsum_epsilon" = "cumsum(epsilon)",
      "conc" = "Concentration")
  ) %>%
  ggplot(aes(x = day, y = value, group = path_number, col = path_number)) +
    geom_line() +
    facet_wrap(~variable, ncol = 1, scales = "free") +
    scale_colour_manual(values = cbpalette) +
    theme_minimal() +
    labs(x = "Day", y = "", col = "Path number", title = "Exponential behaviour")
```

## Sequencing observation models

<!-- * there is additional noise + bias in the seq measurements which I also would model as lognormal but fine to ignore for now -->
See @townes2020review for a nice review of models for count data.

### Dirichlet-multinomial model

The Dirichlet is a probability distribution over proportions. 
$w \sim \text{Dirichlet}(\alpha_1, \ldots, \alpha_K)$ if
$$
p(w; \alpha) = \frac{\Gamma(\sum_k \alpha_k)}{\prod_k \Gamma(\alpha_k)} w_1^{\alpha_1 - 1} w_2^{\alpha_2 - 1} \times \cdots \times w_K^{\alpha_K - 1}
$$

```{r}
dirichlet_draw <- gtools::rdirichlet(n = 1, alpha = K_norm[1, ])
```

### Poisson-lognormal model

The multivariate Poisson-lognormal model (MPLN) [@chiquet2021poisson] is set-up for analysis of an abunance table, an $n \times p$ matrix $Y$ where $Y_{ij}$ is the number of individuals from species $j \leq p$ observed at site $i \leq n$.
"Species" could refer to an operational taxonomic unit (OTU) or amplicon sequence variant (ASV), "site" could refer to a sample or experiment, and "number of individuals" could refer to number of reads.
One model is
$$
Y_{ij} \, | \, Z_{ij} \sim \text{Poisson}(\exp(o_{ij} + Z_{ij})), \\
Z_i \sim \mathcal{N}(\mu_i, \Sigma),
$$
where $Z_i$ are assumed to be independent across sites.
Some moments of $Y$ are given by:
$$
\mathbb{E}(Y_{ij}) = \exp(o_{ij} + \mu_{ij} + \sigma_{jj} / 2) > 0 \\
\mathbb{V}(Y_{ij}) = \mathbb{E}(Y_{ij}) + \mathbb{E}(Y_{ij})^2 (\exp(\sigma_{jj} - 1)) > \mathbb{E}(Y_{ij}) \\
\mathbb{Cov}(Y_{ij}, Y_{ik}) = \mathbb{E}(Y_{ij}) \mathbb{E}(Y_{ik}) (\exp(\sigma_{jk}) - 1)).
$$
Note:
1. Expected count is greater than just adding the exponential of the mean of the latent layer to the expected log abundances because of the logarithmic link function.
2. The model displays over-dispersion as compared with a Poisson model where the variance is the same as the mean, because of the latent noise. (I'm still interested as to how this is different from e.g. a negative-binomial model, though I prefer to use random effects for noise as they are easier to modify and extend to include structure say [amongst other reasons] than likelihoods).
3. "Faithful correlation" in that the correlation and covariance have the same sign. As well, if $\sigma_{jk} = 0$ then $\mathbb{Cov}(Y_{ij}, Y_{ik}) = 0$.

## Bibliography {-}
